Windowed Developer Topic Analysis

### 2009-04-09 lda-paper.tex
big migod edits

What happened in the previous maintenance or development iteration? What were the developers
working on? What was the developer in the next cubicle working on?
What were the common topics and issues dealt with in the current
release of the software? What were the topics of the previous release?
Given the requirements agreed to at the start of the iteration did our
developers work on them? What else did they work on.

Topic analysis, with respect to Software Control Systems (SCSs) like
CVS, Subversion, Git etc., is potentially valuable as it has many uses
for various stake-holders and uses.
Imagine a
development team has agreed to implement some features at the start of
an iteration. By the end of the iteration only some of the agreed upon
features are implemented. The developers swear they were focusing
solely on the features agreed upon. Topic analysis could help by
teasing out the distinct topics or focusses of the developers. Perhaps
the developers had to deal with issues.
Developers might have grouped those orthogonal changes into
one of the features they were working on, thus when asked about how
their time was spent the orthogonal changes would not be covered. This
is what we hope that windowed topic analysis can uncover. If windowed
topic analysis can discover important but overlooked topics, it can
help both developers and managers. Developer can see a
summary of what worked on, and managers can development.

Commits in SCSs are not necessarily consistent, they can have multiple
purposes, thus we hope that these multiple purposes manifest themselves as
topics interleaved within the commits.  Sometimes a single revision to
a file contains multiple bug fixes.  Others have demonstrated that by
using powerful topic analysis techniques like Latent Semantic Indexing
(LSI) or Latent Dirichlet Allocation (LDA) one can separate topics
from these mixed comments. These topics often describe separate ideas
that are interleaved throughout the development. A topic represents a
groups of commit comments which are related to each other by their
content.  In this case a topic is a set of tokens that are related to
each other.

@@

One use of topic analysis is to help us partition a project's
time-line into periods of developer focus. By breaking apart an
iteration into sets of topics and trends (topics which reoccur), we
may be able to recognize the underlying software development process
from these commits. Alternatively we can track down what maintenance was
occurring in the repository at a particular time.

We propose to extend the use of common topic analysis techniques by
executing them over windows of documents over time. We hope by looking
at topic clusters during periods or windows over time we can identify
continuous effort on a select group of topics as well reflect topics
that were local to that period. Figure \ref{fig:lda} is an example of
the kind of visualization we want to automatically generate. We want
to see topics which are unique to a certain time period, displayed
across the time axis. Those topics that reoccur, or occur over a
larger period are plotted continuously. In our example we have
summarized the topics with a single word.

In this paper we explore how topics shift over time in the
source control repository of a project. We analyze commit comments per
revision in each time window and we extract the topics in that time
window. We expect that topics in each window will change over time and
the similarities and differences in these sets of topics will indicate
both developer focus and changes in developer focus.

Our contributions include:
\begin{itemize}
\item Demonstrating the value of windowed topic analysis using trends
\item Multiple visualizations topics and trends over time
\end{itemize}

@@

Some terms we will be using for the rest of the paper include:
message, word distribution, topics and trends. \emph{Messages} are
blocks of text written by developers. In this paper messages will be
the CVS commit comments made when the user commits changes to files in
a repository. \emph{Word distributions} are the summary of a message
by word count. The word distribution will be over all messages looked
at but since most words will not appear in a message, a word
distribution is effectively a word count divided by the message
size. A \emph{topic} is essentially a set of words which form a word
distribution that is unique and independent within the set of
documents in our total corpus. One could think of a topic as the
center distribution of a group of messages. In this paper we often
summarize topics by their top 10 most frequent words in their word
distribution.  A \emph{trend} is a one or more similar topics which
reoccur over time. Trends are important because they are sets of
topics which reoccur over the development of a project.

@@

%XXX explain LDA \cite{944937}

In terms of clustering and finding topic distributions, Latent
Dirichlet Allocation (LDA)~\cite{944937} competes with Latent Semantic
Indexing (LSI), probabilistic Latent Semantic Indexing (pLSI) and
semantic clustering. These tools is used for document modelling,
document clustering and collaborative filtering. What LDA attempts to
do is encode documents as mixture model of topics.  LDA is used in
software engineering
literature~\cite{lukins2008,10.1109/MSR.2007.20,NIPS2007637,1321709}
is to extract topic clusters from documents such as methods, bug
reports, source code and files.

Each of these, LDA, LSI and semantic clustering, extract topic
clusters from the documents that are independent from one another. LDA
doesn't name these clusters but the clusters consist of words whose
relationship is often obvious to someone familiar with the corpus. We
could swap LDA for LSI or semantic clustering and produce similar
results. Our data is posed as documents with word distributions (word
counts per documents) and LDA extracts distinct topics (clusters of
words) from the documents.

@@

In order to infer or associate the expertise of an author with topics
extracted from SCS, Linstead et al proposed an author-source-code model
using LDA~\cite{10.1109/MSR.2007.20,NIPS2007637,1321709}. This model
is essentially an extension of the author-topic (AT)
model~\cite{1036902}, which associated authors and topics extracted
with LDA.

Lukins, Kraft and Etzkorn~\cite{lukins2008} use LDA to help bug
localization by querying for related documents to a bug's topic. They
used LDA to build a topic model and then queried against it using
sample documents. These queries would indicate the proportion of
topics that were associated with the query and thus possible the
related documents.

LSI is related to LDA and has been used to identify topics in software
artifacts for formal concept analysis and concept
location~\cite{1421013,1374321,10.1109/ICPC.2007.13,10.1109/ICPC.2006.17}.
Concept analysis aims to automatically extract concepts from source
code or documents related to the source code.  Concept location is how
high level concepts relate to low level entities like source code, for
instance when fixing a bug how and where do the concepts in the bug
report map back to the source code?  Semantic Clustering has also been
used for similar reasons~\cite{1698774,1566153} as it is similar to
LSI.

Grant et al.~\cite{scottcordy} have used an alternative technique,
rather than LSI and LDA, called Independent Component Analysis to
separate topic signals from software code.

Our technique differ from the previous techniques because we apply LDA
locally to month-long windows of commit comments, where as other
approaches apply LDA once to the entire project.

@@

\section{First Impressions}

In our first exploratory pass we wanted to see if LDA could provide
interesting topics extracted from a real system. We took the
repository of MySQL 3.23, extracted the commits, grouped them by 30
day non-overlapping windows then applied LDA to each of these
windows. We asked LDA to make 20 topics per window and we would pay
attention to the top 10 most frequent words in that topic.  We found
there were common words across topic clusters, such as diffs, annotate
and history, these words probably should be viewed as stop
words. There were notable transitional topics such as in the first
window the word ``bitkeeper'' appears (probably MySQL they adopted
Bitkeeper for their source control system) yet in the following
windows there were no mentions of Bitkeeper. ``RENAME'' also only
appeared once in the first window and never again. Per each topic we
tried to name the purpose of the topic, we found that with little
familiarity with the code base that naming the topic was not difficult.

A sub-sampling of the notable topic words is displayed in Table
\ref{tab:portability}.  To name the topics we picked out terms which
seem to summarize those topics.  After viewing these topic is was
obvious that we would have to combine try to track the evolution of
topics over the windows or rank topics by similarity, thus if they
were similar enough we joined them into a trend. Figure \ref{fig:lda}
displays a sample plot of the extracted topics in table
\ref{tab:portability}.

@@

\begin{figure*}
  \centering
  \includegraphics[width=0.9\textwidth]{lda}
  \caption{Example of topics extracted from MySQL 3.23. This is the kind of plot we eventually want to produce: named topics and topic trends. The horizontal axis is time in months and the vertical axis used to stack topics that occur at the same time. Longer topics are topics which reoccur in adjacent window. Colors are arbitrary.}
  \label{fig:lda}
\end{figure*}

@@

Our methodology is to first extract the commits from repositories such
as CVS or Bitkeeper. Then we extract the revisions and commits,
grouping the revisions into commits if necessary. We extract the
commit log message of each commit, count the occurrence of each word in
the message, remove stop words, and then produce word distributions
for each message.

\subsection{Extract Repos}

We extracted the repositories with software such as rsync, CVSSuck,
\texttt{softChange} and bt2csv.  \texttt{softChange} provided a
general schema for storing revisions. CVSSuck and rsync mirrored CVS
repositories while bt2csv mirrored web accessible bitkeeper
repositories.

@@

The document we are analyzing are the changelog comments, these are
the comments added when a revision to the source code is committed.
Documents are converted into word count distributions. These
distributions are can be normalized by by the size of the
message. After all messages are processes the distributions are
extended to include all words in all distributions (each distribution
is extended with word bins of count 0).

@@

Given all messages, we group the messages into windows. We could use
overlapping windows, but in this paper we use non-overlapping windows
of a set time unit.  Windowing by time allows for many levels of
granularity. We usually used one month as the length of our time
windows, while we could use different granularities for the study we
think that a month is a sizable unit of development (4 weeks of
development).

@@

Once we have our data windowed, we apply our Topic Analysis tool to
each window and extract the top $N$ topic clusters, we used $20$
topics in our case study. Our Topic Analysis tool is an implementation
of LDA, but we could've used other similar tools like LSI.

This is a very slow step as LDA takes a long time to run, and takes
even longer since we run it once per window.

@@

Once we have our topics extracted for each window , we analyze them
and look for repeating topics. We do this by taking the top 10 most
common words and then comparing them.

We then abstract these topics by comparing them to each other.  Each
topic is compared to each other topic in the system, given a certain
threshold of top 10 similarity, like 8/10 words. Then given this topic
similarity matrix, we find the transitive closure of similar topics,
that is we fill flood along the similarity arcs until we have
partitioned the topics into clusters of similar topics. Figure
\ref{fig:closure} provides an example of clustering topics by the
topic similarity. These clusters of topics are called trends. A trend
is probably interesting and valuable if there is more than 1 topic in
the trend.

This technique has weaknesses in that nodes a few neighbors away in
similarity might not share any similar words.  We do this because we
want to be able to color or highlight topics which are related to each
other and track them throughout time.

@@

We have come up with multiple ways to visualize these topics and
trends.  For all of these plots if we trends which have continuous
segments we plot those segments as joined across time.  We have a
compressed stacked trend view, shown in Figure \ref{fig:topicsmear}
and Figure \ref{fig:zoomedsmear}, which displays trends arbitrarily
across the y axis but are placed where they occur on the time-based x
axis. Another plot we have is the trend-time line plot, shown in
Figure \ref{fig:timeline}, each trend gets it own distinct-y value ,
while the x-axis is time; these topics are then each plotted on their
own line across time as they occur. Our final plot is the histogram
plot, shown in Figure \ref{fig:histogram} where we plot each trend on
its own line but stack up the segments of the trend, much like a plot
of counts or a histogram .

The first visualization is a compressed view (Figure \ref{fig:topicsmear}
and Figure \ref{fig:zoomedsmear}), we try to sink the
larger trends to the bottom, once the larger trends are stacked, we
fill in the gaps with the smaller trends in the same window and then
stack the smaller trends on top. Although there is a chance that gaps
will be left due to non-optimal stacking in practice there are lot of
small trends (90\% of all trends contain 1 topic) which fill in these
gaps quickly. The trend segments that share the same color are the
same trend, colors are randomly chosen for each trend, so color
similarity is meaningless, only exact color matching matters.
The compressed view is nice because it shows all of the
information and the repeating continuous trends are easy to pick out,
unfortunately repeating discontinuous trends are harder to spot.

The second visualization is the trend-time line plot (Figure
\ref{fig:timeline}), it displays repeating trends more clearly by
dedicating a horizontal line for trend segments dedicated to one
trend. This means that if a trend contains discontinuous segments the
segments appear on the same line. One down side to this view is the
least common trends need to be pruned otherwise it'd take too many
horizontal lines to plot them.

The third visualization is the histogram plot (Figure
\ref{fig:histogram}) which is a lot like the time-trend plot except
the trends are plotted together by stacking to their row so time
information is lost and the trends are ordered by the number of topics
in the trend. The histogram plot shows the count of instances of a
trend and thus indicates which trends occur the most. Unfortunately
due to the large number of topics given allotted space it is often
best to crop off the trends with only one topic, otherwise the tail is
long.

\begin{figure*}
  \centering
  \includegraphics[width=1.0\textwidth]{fixed-time-smear-plot-scaled}
  \caption{Compact Trend Plot Topics per Month for MaxDB 7.500. The
    X-axis is time in months, the Y axis is used to stack topics
    occurring at the same time. Trends which are continuous are
    plotted as continuous blocks. The top 10 words in the topics are
    joined and embedded in box representing the topics.}
  \label{fig:topicsmear}
\end{figure*}

@@

Our analysis tools consisted of: Hiraldo-Grok, an OCaml based spin off
of Grok used for answering queries; Gnuplot, a graph plotting package;
lda-c, a LDA package implemented by Biel; and our topic plotter,
implemented in Haskell.

@@

We applied our tools and visualizations to the repositories of
multiple database systems.

@@

PostgreSQL did not have many trends with two or more topics, but it
did have one large trend which stretched across most of its
existence. Much of the large trend consisted of stop words but it did
include topics which had words such as patch, fix, update.

The second largest topic was interesting because it dealt changes from
1998, from developers using Central Europe Time (CEST), the two
notable tokens were 1998 and CEST. These changes seem to be patches
submitted by one particular developer who used that timezone.

The third largest trend seemed to be about copying and moving files
between branches as it referenced the old branch. This is a side
effect of the programmers manually recording such a merge across
branches because CVS cannot.

Once stop words were removed the some of the purposes became more
clear. The large trend disappeared, the largest trend became bug fixes
submitted by Dal Zotto which is a short trend.

% Large Trends is not the point!!!
% That's what normal LDA gives us

But large trends are not the point, by investigating windowed trends
we want more local information about trends being worked on.

@@

We tracked Firebird from August 2000 to January 2006.
Our first attempt with Firebird resulted in a set of trends that were
dominated by one large trend containing tokens which were mostly stop
words such as ``the to a for of is it'' and references to one author
\texttt{carlosga05}.

The second largest trend related to branching and possibly merging, it
contained language such as ``was added on branch'' and contained
references to tags and branches.

The third largest trend contained reference to the changelog, the
build system, architecture and platform support. It specifically
mentioned tokens such as ``Solaris port x86''.

@@

Topics that were not trends that were interesting were external bug
fixes submitted to the project, the developers would write thank yous
in their changelog comments: ``thanks Bill Lam''. Other interesting
topics include tokens and topics such as: ``compiler workarounds'', ``nightly updates'', packets and MSVC++,

@@

The plots we produced of MaxDB 7.500 were interesting, there was a
period where no real development occurred and thus there were no topics or trends
whatsoever. We have two periods to evaluate for MaxDB 7.500, the first
period from June 2004 to jan 2005, and the second period from June
2005 to June 2006.

The largest common trend has references to build system files like
SYSDD.punix and MONITOR.punix etc. Other tokens mentioned are Sutcheck
v1.09, the prefix SUT stands for Storable Unit Type. Sutcheck would
also automate perforce check-ins.

The second largest common trend seems to be a side effect of an
automated checkin which is annotated as ``implicit checkin''. These
were checkins that were produced when importing changes from an
external Perforce repository.

@@

The previous work which leverage LSI or LDA usually extracted a
certain number of topics and tracked them over time. This meant a
static set of topics was tracked across the whole development history
of a project. Thus we really had 20 trends.

We attempted the same kind of analysis, as shown in
Figure~\ref{fig:statictopics} we extracted 20 topics and then plotted
the number of messages per month which related to that topic. What we
found was that often one topic would dominate the results, while
messages related to other topics didn't appear that often.

This technique seems useful if the 20 extracted topics are relevant
and useful, that said one weakness of this approach is that documents
for a topic might only occur in one time window, yet we're tracking
messages that could be related to it over the entire lifetime of the
project.

@@

In this study we are explicitly trusting that the programmers annotate
their changes with relevant information. We rely on the descriptions
they provide. If the language of checkin comments was automated we
would only be analyzing that.

@@

Each project seemed to have their own kind of stop words, we did not
remove them, but perhaps dropping some of these stop words would aid
the clarity of such topic clusters, alternatively it might cause our
topic similarity plots to be fundamentally different. Perhaps our
methods of suggesting that topics are similar is not appropriate.

The number of commits per month is inconsistent some months have a lot
of changes where as some months have almost none.

%   1.   validation
%   2.   LDA is questionable
%   3.   blackbox problem
%   4.   not in the token
%   5.   multiple X token
%9.    Future Work

@@

Even with our liberal topic similarity metrics which produced both
long and short trends, we showed that there are only a few trends in a
repository which re-appear. Since so few trends re-appear and so many
trends appear only once this suggests that global topic analysis might
be ignoring locally unique topics.

@@

We presented multiple visualizations which focused on different
aspects of the data: temporality of trends, trend size, and a
condensed overall trend view. The condensed trend view shows much more
information than the views that global analysis could show.

@@

One avenue of future work we think is quite important is automatic
cluster labelling. Given a word distribution we should be able to
figure out a word or term which best describes that distribution
automatically.
Another avenue we wish to investigate the is value of applying
static community generated software taxonomies to the naming or
grouping of clusters

>>>>

What happened in the previous maintenance or development iteration? What were the developers
working on? What was the developer in the next cubicle working on?
Which common topics and issues were dealt with in the current
release of the software? What were the topics of the previous release?
Given the requirements agreed to at the start of the iteration did our
developers work on them? What else did they work on?

Topic analysis, with respect to Software Control Systems (SCSs) such
as CVS, Subversion, Git etc., is potentially valuable as it has many
uses for various stake-holders and uses.  Imagine a development team
has agreed to implement some features at the start of an iteration. By
the end of the iteration only some of the agreed upon features are
implemented. The developers insist they were focusing solely on the
features agreed upon, but what actually happened? Topic analysis could
help by teasing out the distinct topics or focusses of the
developers. Perhaps the developers had to deal with issues.
Developers might have grouped those orthogonal changes into one of the
features they were working on, thus when asked about how their time
was spent the orthogonal changes would not be covered. This is the kind of underlying story that we
hope that windowed topic analysis can uncover. If windowed topic
analysis can discover important but overlooked topics, it can help
both developers and managers. Developer can see a summary of what
worked on, and managers can development.

Commits in SCSs are not necessarily consistent, they can have multiple
purposes, thus we hope that these multiple purposes manifest themselves as
topics interleaved within the commits.   Others have demonstrated that by
using powerful topic analysis techniques, such as Latent Semantic Indexing
(LSI) and Latent Dirichlet Allocation (LDA), can aid in separating topics
from these mixed comments. These topics often describe distinct ideas
that are interleaved throughout the development. A topic represents a
groups of commit comments that are related to each other by their
content.  In this case a topic is a set of tokens.

@@

One use of topic analysis is to help us partition a project's
time-line into periods of developer focus. By breaking apart an
iteration into sets of topics and trends (topics which recur), we may
be able to recognize the underlying software development process from
these commits. Alternatively, we can determine what particular
maintenance task was being performed are a given time.

We propose to extend the use of common topic analysis techniques by
executing them over windows of documents over time. We hope by looking
at topic clusters during periods or windows over time we can identify
continuous effort on a select group of topics as well reflect topics
that were local to that period. Figure \ref{fig:lda} is an example of
the kind of visualization we want to automatically generate. We want
to see topics that are unique to a certain time period, displayed
across the time axis. Those topics that recur, or occur over a
larger period are plotted continuously. In our example we have
summarized each topic with a single word.

In this paper we explore how topics shift over time in the source
control repository of a project, using several open source database
systems as examples. We analyze commit comments per revision in each
time window and we extract the topics in that time window. We expect
that topics in each window will change over time and the similarities
and differences in these sets of topics will indicate both developer
focus and changes in developer focus.

Our contributions include:
\begin{itemize}
\item Demonstrating the value of windowed topic analysis using trends
\item Multiple visualizations topics and trends over time
\item An exploratory case study using these techniques on several open source database systems.
\end{itemize}

@@

We now define some terms that we will use in the rest of the paper: A
\emph{message} is a block of text written by developers. In this
paper, messages will be the CVS commit comments made when the user
commits changes to files in a repository. A \emph{word distribution}
is the summary of a message by word count. The word distribution will
be over all messages looked at but since most words will not appear in
a message, a word distribution is effectively a word count divided by
the message size. A \emph{topic} is essentially a set of words that
form a word distribution that is unique and independent within the set
of documents in our total corpus. One could think of a topic as the
center distribution of a group of messages. In this paper we often
summarize topics by their top 10 most frequent words in their word
distribution.  A \emph{trend} is a one or more similar topics which
recur over time.  Trends are particularly important, as they indicate
long-lived and recurring topics that may provide key insights into the
development of the project.

@@

%XXX explain LDA \cite{944937}
%XXX What has pLSI in it?
In terms of clustering and finding topic distributions, Latent
Dirichlet Allocation (LDA)~\cite{944937} competes with Latent Semantic
Indexing (LSI)~\cite{1421013,1374321,10.1109/ICPC.2007.13,10.1109/ICPC.2006.17}, probabilistic Latent Semantic Indexing (pLSI) and
semantic clustering~\cite{1698774,1566153}. These tools are used for document modelling,
document clustering and collaborative filtering. What LDA attempts to
do is encode documents as mixture model of topics.  LDA is used in
software engineering
literature~\cite{lukins2008,10.1109/MSR.2007.20,NIPS2007637,1321709}
is to extract topic clusters from documents such as methods, bug
reports, source code and files.

LDA, LSI and semantic clustering extract topic
clusters from the documents that are independent from one another. LDA
does not name these clusters but the clusters consist of words whose
relationship is often obvious to someone familiar with the corpus. We
could swap LDA for LSI or semantic clustering and produce similar
results. Our data is posed as documents with word distributions (word
counts per documents) and LDA extracts distinct topics (clusters of
words) from the documents.

@@

In order to infer or associate the expertise of an author with topics
extracted from SCS, Linstead et al. proposed an author-source-code model
using LDA~\cite{10.1109/MSR.2007.20,NIPS2007637,1321709}. This model
is essentially an extension of the author-topic (AT)
model~\cite{1036902}, which associated authors and topics extracted
with LDA.

Lukins, Kraft and Etzkorn~\cite{lukins2008} use LDA to help bug
localization by querying for documents that are related to a bug's topic. They
used LDA to build a topic model and then queried against it using
sample documents. These queries would indicate the proportion of
topics that were associated with the query and thus possible the
related documents.

LSI is related to LDA and has been used to identify topics in software
artifacts for formal concept analysis and concept
location~\cite{1421013,1374321,10.1109/ICPC.2007.13,10.1109/ICPC.2006.17}.
Concept analysis aims to automatically extract concepts from source
code or documents that are related to the source code.  Concept location concerns how
high level concepts relate to low level entities such as source code, for
example, when fixing a bug how and where do the concepts in the bug
report map back to the source code?  Semantic Clustering has also been
used for similar purposes~\cite{1698774,1566153} as it is similar to
LSI.

Grant et al.~\cite{scottcordy} have used an alternative technique,
 called Independent Component Analysis to
separate topic signals from software code.

Our technique differs from the previous techniques because we apply
LDA locally to month-long windows of commit comments, where as other
approaches apply LDA once to the entire project. This allows us to
examine the time-based windowing of topics over its development
history.

@@

\section{Preliminary Case Study}

In our first exploratory pass we wanted to see if LDA could provide
interesting topics extracted from a real system. We took the
repository of MySQL 3.23, extracted the commits, grouped them by 30
day non-overlapping windows, and then applied LDA to each of these
windows. We asked LDA to make 20 topics per window and we then
carefully examined the top 10 most frequent words in that topic.  We
found there were common words across topic clusters, such as
``diffs'', ``annotate'' and ``history'', these words probably should
be viewed as stop words. There were notable transitional topics such
as in the first window the word ``bitkeeper'' appears because MySQL
adopted Bitkeeper for their source control system yet in the following
windows there were no mentions of Bitkeeper. ``RENAME'' also only
appeared once in the first window and never again. Per each topic we
tried to identify the purpose of the topic; to our surprise we found
that even with onlyy a little familiarity with the code base that
naming the topic was straightforward.

A sub-sampling of the notable topic words is displayed in Table
\ref{tab:portability}.  To name the topics we selected terms that
seemed to best summarize those topics.  After viewing these topics, it was
obvious that we would have to combine try to track the evolution of
topics over the windows or rank topics by similarity, thus if they
were similar enough we joined them into a trend. Figure \ref{fig:lda}
displays a sample plot of the extracted topics in table
\ref{tab:portability}.

@@

\begin{figure*}
  \centering
  \includegraphics[width=0.9\textwidth]{lda}
  \caption{Example of topics extracted from MySQL 3.23. This is the kind of plot we eventually want to produce: named topics and topic trends. The horizontal axis is time in months and the vertical axis used to stack topics that occur at the same time. Longer topics are topics which recur in adjacent window. Colors are arbitrary.}
  \label{fig:lda}
\end{figure*}

@@

Our methodology is to first extract the commits from SCS repositories such
as CVS or Bitkeeper. Then we extract the revisions and commits,
grouping the revisions into commits if necessary. We extract the
commit log message of each commit, count the occurrence of each word in
the message, remove stop words, and then produce word distributions
for each message.

\subsection{Extract Repositories}

We mirrored the repositories and their revisions using software such
as rsync, CVSSuck, \texttt{softChange} and bt2csv.
\texttt{softChange} provided a general schema for storing
revisions. CVSSuck and rsync mirrored CVS repositories while bt2csv
mirrored web accessible bitkeeper repositories.

@@

The document we are analyzing are the changelog comments i.e., the comments that are added  when a revision to the source code is committed.
Documents are converted into word count distributions. These
distributions, which are then normalized by by the size of the
message. After all messages are processed the distributions are
extended to include all words in all distributions (each distribution
is extended with word bins of count 0).

@@

Given all messages, we group the messages into windows. We could use
overlapping windows, but in this paper we use non-overlapping windows
of a set time unit because it simplifies analysis.  Windowing by time
allows for many levels of granularity. We usually used one month as
the length of our time windows, while we could use different
granularities for the study we think that a month is a sizable unit of
development, 4 weeks of development, which we consider to be fine
enough to show meaningful shifts but coarse enough not to show too
much detail.

@@

Once we have our data windowed, we apply our Topic Analysis tool to
each window and extract the top $N$ topic clusters, we used $20$
topics in our case study. Our Topic Analysis tool is an implementation
of LDA, but we could have used other similar tools like LSI, but
previous showed more promising topic analysis with LDA.

This is a very slow step as LDA takes a long time to run, and takes
even longer since we run it once per window. We note that this step is
a slow one, as executing even one instance of LDA involves significant
computation, and we perform LDA once per window.

@@

Once we have our topics extracted for each window, we analyze them and
look for repeating topics. We find repeating topics by comparing the
10 most common words of each topic.

We then abstract these topics by comparing them to each other.  Each
topic is compared to each other topic in the system, given a certain
threshold of top 10 similarity, such as 8/10 words. Then given this topic
similarity matrix, we find the transitive closure of similar topics;
that is, we fill flood along the similarity arcs until we have
partitioned the topics into clusters of similar topics. Figure
\ref{fig:closure} provides an example of clustering topics by the
topic similarity. These clusters of topics are called trends. A trend
is probably interesting and valuable if there is more than one topic in
the trend.

This technique has weaknesses in that nodes that are  a few neighbors away in
similarity might not share any similar words.  We do this because we
want to be able to color or highlight topics that are related to each
other and track them throughout time.

@@

We have devised several ways to visualize these topics and
trends.  For all of these plots if we find trends that have continuous
segments, then we plot those segments as joined across time.  We have a
compressed stacked trend view, shown in Figure \ref{fig:topicsmear}
and Figure \ref{fig:zoomedsmear}, 
which displays trends as they occur
on the time-based x-axis (placement along the y-axis is arbitrary).
 Another plot we have is the trend-time line plot, shown in
Figure \ref{fig:trendtimeline}, each trend gets it own distinct y-value ,
while the x-axis is time; these topics are then each plotted on their
own line across time as they occur. Our final plot is the histogram
plot, shown in Figure \ref{fig:histogram} where we plot each trend on
its own line but stack up the segments of the trend, much like a plot
of counts or a histogram.

The first visualization is a compressed view
(Figure\ref{fig:topicsmear} and Figure \ref{fig:zoomedsmear}) where we
try to sink the larger trends to the bottom.  Once the larger trends
are stacked we fill in the gaps with the smaller trends in the same
window, and then stack the smaller trends on top.  Although there is a
chance that gaps will be left due to non-optimal stacking in practice
there are lot of small trends (90\% of all trends contain 1 topic)
which fill in these gaps quickly.  Different instances of the same
trend share the same colour; apart from that, the color of trends is
randomly chosen and color similarity is not meaningful.
The compressed view is convienant because it shows all of the
information and the repeating continuous trends are easy to pick out,
although repeating discontinuous trends are harder to spot.

The second visualization is the trend-time line plot (Figure
\ref{fig:trendtimeline}), it displays repeating trends more clearly by
dedicating a horizontal line for trend segments dedicated to one
trend. This means that if a trend contains discontinuous segments then the
segments appear on the same line. 
However, the least common trends need to be pruned away or the view will be cluttered with a lot of noise.

The third visualization is the histogram plot (Figure
\ref{fig:histogram}) which superficially resembles the time-trend
plot.  However, in this view the trends are plotted together by
stacking to their row so time information is lost, and the trends are
ordered by the number of topics in the trend.  The histogram plot
shows the count of instances of a trend and thus indicates which
trends occur the most. Unfortunately due to the large number of topics
given allotted space it is often best to crop off the trends with only
one topic, otherwise the tail is long.

%XXX Describe something useful like a trend
\begin{figure*}
  \centering
  \includegraphics[width=1.0\textwidth]{fixed-time-smear-plot-scaled}
  \caption{Compact Trend Plot Topics per Month for MaxDB 7.500. The
    X-axis is time in months, the Y axis is used to stack topics
    occurring at the same time. Trends that are continuous are
    plotted as continuous blocks. The top 10 words in the topics are
    joined and embedded in box representing the topics.}
  \label{fig:topicsmear}
\end{figure*}

@@

Our analysis tools consisted of: Hiraldo-Grok, an OCaml-based variant
of the Grok query language; Gnuplot, a graph plotting package; lda-c,
a LDA package implemented by Biel; and our topic plotter, implemented
in Haskell.

@@

We applied our tools and visualizations to the repositories of several
open source database systems: PostgreSQL, MaxDB and Firebird.

@@

When we examined PostgreSQL, we did not find many trends with two or more topics; however, we did find one large trend which stretched across most of its existence.
 Much of the large trend consisted of stop words but it did
include topics which had non-stop words such as \emph{patch}, \emph{fix}, \emph{update}.

The second largest topic was interesting because it concerned changes from
1998, from developers using Central Europe Time (CEST), where the two
notable tokens were 1998 and CEST. These changes appear to be patches
submitted by one particular developer who lived in that timezone.

The third largest trend we found appeared to concerned with copying
and moving files between branches as it referenced the old branch.
branches.  We inferred this trend based on the fact that the old
version identifier was referenced, and we know this this is common
practice when performing a branch merge, since CVS cannot record such
information directly.

Once stop words were removed the some of the purposes became more
clear. The large trend disappeared, the largest trend became bug fixes
submitted by a developer named Dal Zotto which is a short trend.

%XXX REDO POSTEGRESQL?
% Large Trends is not the point!!!
% That's what normal LDA gives us

%But large trends are not the point, by investigating windowed trends
%we want more local information about trends being worked on.

@@

We tracked Firebird from August 2000 to January 2006.
Our first attempt with Firebird resulted in a set of trends that were
dominated by one large trend containing tokens that were mostly stop
words such as ``the to a for of is it'' and references to one author
\texttt{carlosga05}.

The second largest trend related to branching and possibly merging; it
contained language such as ``was added on branch'' and contained
references to tags and branches.

The third largest trend contained references to the changelog, the
build system, architecture and platform support. It specifically
mentioned tokens such as ``Solaris port x86''.

@@

Topics that were not trends but appeared to be  interesting were mostly external bug
fixes submitted to the project.
 In these cases, the developers would express thanks in their changelog comments, such as ``Thanks, Bill Lam''.  Other easily discernible topics included 
 tokens and topics such as: ``compiler workarounds'', ``nightly updates'', packets and MSVC++,

@@

%XXX Migod todo
% I think that you should enlarge this discussion of MaxDB (at the
% expense of the other systems) and make careful references to how the
% diagrams give you extra info about the evolution of the system (in
% terms of topics).

% We are trying to argue that (a) the approach is more informative that
% the naive static approach and (b) the diagrams are useful.

The plots we produced of MaxDB 7.500 were 
unlike those of the other systems, as 
 there was a
period where no real development occurred and thus there were no topics or trends
whatsoever. We have two periods to evaluate for MaxDB 7.500, the first
period from June 2004 to jan 2005, and the second period from June
2005 to June 2006.

The largest common trend has references to build system files like
SYSDD.punix and MONITOR.punix etc. Other tokens mentioned are Sutcheck
v1.09, the prefix SUT stands for Storable Unit Type. Sutcheck would
also automate check-ins using a Perforce SCS tool.

The second largest common trend seems to be a side effect of an
automated check-in which is annotated as ``implicit check-in''. These
were check-ins that were produced when importing changes from an
external Perforce repository.

@@

%XXX Migod notes This section is perhaps the most important of the
% whole paper.  You need expand it.  Maybe you should move the first
% two paras of the conclusions in here too.

% The thing that is most disappointing is that the argument is entire qualitative.  Can you give examples to make this sound less hypothetical?

% Can you add numbers in here to make the argument stronger?

Previous work on topic analysis that employed LSI and LDA typically
extracted a certain number of topics and tracked them over time. This
means that a single, static set of topics was tracked across the whole
development history of a project.

We attempted the same kind of analysis, as shown in
Figure~\ref{fig:statictopics}. We extracted 20 topics and then plotted
the number of messages per month that were related to that topic. We
found was that one topic would often dominate the results, while
messages related to other topics did not appear very often.
% This technique seems useful if the 20 extracted topics are relevant
% and useful, that said one weakness of this approach is that documents
% for a topic might only occur in one time window, yet we are tracking
% messages that could be related to it over the entire lifetime of the
% project.
This approach seems reasonable if most of the extracted topics are of
broad interest during most of the development process.  However, it
may be that some topics are of strong interest but only briefly; in
such a case, a window-based model gives a much stronger indication of
the fleeting importance of such topics, and can help to put such a
topic into its proper context.

@@

In this study we are explicitly trusting that the programmers annotate
their changes with relevant information. We rely on the descriptions
they provide. If the language of check-in comments was automated we
would only be analyzing that.

@@

Each project seemed to have their own kind of stop words, we did not
remove them, but perhaps dropping some of these stop words would aid
the clarity of such topic cluster; alternatively it might cause our
topic similarity plots to be fundamentally different.

The number of
commits per month is inconsistent as some months have a lot of changes
while others months have almost none.
%   1.   validation
%   2.   LDA is questionable
%   3.   blackbox problem
%   4.   not in the token
%   5.   multiple X token
%9.    Future Work

@@

%XXX migod comments:

% The first two paras here are good stuff, but I think it belongs in
% section 5 at the end, as a summary of the utility of the approach.
% But can you add more meat to it in terms of numbers?

% In the conclusions you can simply restate what we did and what we
% learned without doing the arguing here.

% The third para is a good one for the conclusions, tho there need to
% be more detail.

Even with our liberal topic similarity metrics that produced both
long and short trends, we showed that there are only a few trends in a
repository which recur. Since so few trends recur and so many
trends appear only once this suggests that global topic analysis might
be ignoring locally unique topics.

@@

We presented multiple visualizations that focused on different
aspects of the data: temporality of trends, trend size, and a
condensed overall trend view. The condensed trend view shows much more
information than the views that global analysis could show.

@@

One avenue of future work we think is quite important is automatic
cluster labelling. Given a word distribution we should be able to
automatically select a word or term that best describes that distribution.
Another avenue we wish to investigate the is value of applying
static community-generated software taxonomies to the naming or
grouping of clusters
